Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-07 paper/paper.tex
I moved stuff in the background:

% % % \subsubsection{Histomorphism}

% % % The definition of $\lambda$ is:
% % % \begin{align*}
% % %   & \lambda : GG_{\infty} A \to G_{\infty} GA\\
% % %   & \lambda = \fmap{(\fmap{\head})}\comp \anacofree{\fmap{\tail}}
% % % \end{align*}
% % % For example, for $GX = 1 + X$ we have to define
% % % \begin{align*}
% % %   & \lambda : 1 + G_{\infty} A \to G_{\infty} (1 + A)\\
% % %   & \lambda(\inl) = \pair{\inl}{\lambda(\inl)}\\
% % %   & \lambda(\inr(table)) = \pair{\inr (\head(table))}{\lambda (\tail (table))}
% % % \end{align*}

% % % % \begin{align*}
% % % %   & \operator{lookup} :: \Nat \to G_{\infty} \R \to \operator{Maybe}\;\R\\
% % % %   & \operator{lookup}\; \Zero       (\Cons{a}{\text{\textunderscore}})     = \Just{a}\\
% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Zero})                      = \Nothing\\
% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Succ table})                = \operator{lookup}\; n  table
% % % % \end{align*}

% % % % Defining the algebra
% % % % \begin{align*}
% % % %   & \operator{alg} :: D (G_{\infty} \R) \to \R\\
% % % %   & \operator{alg}\; Zero           = 0\\
% % % %   & \operator{alg}\; (Succ{table}) = maximum_{0} [v + \operator{lookup}\; table (i - w) | (w,v) \leftarrow wvs, w \le i]
% % % % \end{align*}

\subsection{Functors and Containers}
\label{sec:functors}
Functors are functions \coq{F : Type -> Type} which additionally have a
map
\begin{coqcode}
fmap : forall A B, (A -> B) -> F A -> F B
\end{coqcode}
witnessing the idea that a
functor represents a container for abstract data that can be manipulated without
changing the outer structure. For example, the type \coq{List : Type -> Type} is
a functor and \coq{List A} is the type of lists over elements of type \coq{A}
with the obvious \coq{fmap} function recursively traversing a list and applying
the function \coq{A -> B} for each element. Additionally, the type of
lists \coq{List A} arises as the least fixed-point of the
functor \coq{F X = unit + (A * X) }.

In general not every functor has a fixed-point, and it is not possible to build
the fixed-point of a functor such as \coq{F} in Coq due to not being strictly
positive.  Due to this, we are going to restrict to \emph{polynomial functors}
as represented by containers. A container is defined by a type of \emph{shapes}
and a \emph{family of position types} indexed by shapes.
\begin{coqcode}
Context (Shape : Type) (Pos : Shape -> Type).
\end{coqcode}
For the functor \coq{F} we introduced in the previous paragraph we would
define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
the data type and one of these contains a piece of data of type \coq{A}.
Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
and  \coq{Pos (inr a) = unit} indicating that the first constructor does not
have any type variables and the second constructor has one type variable.

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
Recursion schemes provide an abstract way to consume and generate data. We now
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
these notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:rec-coalgebras}).

@@

\section{Extraction}
\label{sec:examples}
We go in this section through a series of case studies of various recursive
algorithms.  We show how they can be encoded in our framework, how can we do
program calculation techniques for optimisation, and how can they be extracted
to idiomatic OCaml code. Our examples are the Quicksort and Mergesort
algorithms (Section~\ref{sec:sorting}), dynammic programming and Knapsack
(Section~\ref{sec:knap}), and examples of the shortcut deforestation
optimisation in our framework (Section~\ref{sec:shortcut}).
%The steps are the following:
%\begin{enumerate}
%  \item Write a specification as a morphism \coq{f : A ~> B}.
%  \item Use Coq's generalised rewriting and the laws of morphisms to 
%    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
%  \item Project the function \coq{h : A -> B} from \coq{g}
%    (i.e. \coq{h = app g}).
%  \item Extract \coq{h} to OCaml.
%\end{enumerate}
%Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
%\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
%proofs from the code, and inline our definitions. To ease the application of
%these steps, we provide the following helper type:
%\begin{coqcode}
%  Record Ext (f : A ~> B) := 
%    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
%\end{coqcode}
%Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
%since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
%will be the function \coq{target} that is extensionally equal to \coq{f}.  We
%use a shortcut tactic, \coq{calculate}, that is a synonym to 
%\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
%\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
%complete the definition with \coq{reflexivity}, and this will instantiate the
%metavariable \coq{?Goal0} to our desired function. We will use this extensively
%in our examples.
%\mvol{This is a very dry overview that doesn't need to be here. We can just jump
%  right into the examples.}
%
%\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
%  be used for a little overview of the rest of the section, like: ``We will go over a series
%  of case studies of various recursive algorithms, showing how they can be encoded
%  in our framework and extracted to idiomatic OCaml code.''}
%
%\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
%  Each of these is written like a case study, with a particular program worked out all the way
%  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
%  a sentence at the beginning of each section re-iterating this to make the whole thing
%  flow more smoothly.}

>>>>

% % % \subsubsection{Histomorphism}
% % % The definition of $\lambda$ is:
% % % \begin{align*}
% % %   & \lambda : GG_{\infty} A \to G_{\infty} GA\\
% % %   & \lambda = \fmap{(\fmap{\head})}\comp \anacofree{\fmap{\tail}}
% % % \end{align*}
% % % For example, for $GX = 1 + X$ we have to define
% % % \begin{align*}
% % %   & \lambda : 1 + G_{\infty} A \to G_{\infty} (1 + A)\\
% % %   & \lambda(\inl) = \pair{\inl}{\lambda(\inl)}\\
% % %   & \lambda(\inr(table)) = \pair{\inr (\head(table))}{\lambda (\tail (table))}
% % % \end{align*}
% % % % \begin{align*}
% % % %   & \operator{lookup} :: \Nat \to G_{\infty} \R \to \operator{Maybe}\;\R\\
% % % %   & \operator{lookup}\; \Zero       (\Cons{a}{\text{\textunderscore}})     = \Just{a}\\
% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Zero})                      = \Nothing\\
% % % %   & \operator{lookup}\; (\Succ{n})  (\Cons{a}{\Succ table})                = \operator{lookup}\; n  table
% % % % \end{align*}
% % % % Defining the algebra
% % % % \begin{align*}
% % % %   & \operator{alg} :: D (G_{\infty} \R) \to \R\\
% % % %   & \operator{alg}\; Zero           = 0\\
% % % %   & \operator{alg}\; (Succ{table}) = maximum_{0} [v + \operator{lookup}\; table (i - w) | (w,v) \leftarrow wvs, w \le i]
% % % % \end{align*}
\section{Recursion Schemes}
\subsection{Functors}
Functors are functions \coq{F : Type -> Type} which additionally have a
map
\begin{coqcode}
fmap : forall A B, (A -> B) -> F A -> F B
\end{coqcode}
witnessing the idea that a functor represents a container for abstract data that
can be manipulated without changing the outer structure. For example, the type
\coq{List : Type -> Type} is a functor and \coq{List A} is the type of lists
over elements of type \coq{A} with the obvious \coq{fmap} function recursively
traversing a list and applying the function \coq{A -> B} for each element.
Additionally, the type of lists \coq{List A} arises as the least fixed-point of
the functor \coq{F X = unit + (A * X) }.

In general not every functor has a fixed-point, and it is not possible to build
the fixed-point of a functor such as \coq{F} in Coq due to not being strictly
positive. Due to this, we are going to restrict to \emph{polynomial functors}
as represented by containers.
\subsection{Algebras and Coalgebras}
\subsection{Hylomorphisms}
Recursion schemes provide an abstract way to consume and generate data. We now
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together these
notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:rec-coalgebras}).

@@

\subsection{Functors and Containers}
\label{sec:functors}
[CUT]
A container is defined by a type of \emph{shapes}
and a \emph{family of position types} indexed by shapes.
\begin{coqcode}
Context (Shape : Type) (Pos : Shape -> Type).
\end{coqcode}
For the functor \coq{F} we introduced in the previous paragraph we would
define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
the data type and one of these contains a piece of data of type \coq{A}.
Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
and  \coq{Pos (inr a) = unit} indicating that the first constructor does not
have any type variables and the second constructor has one type variable.

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
[MOVED to background]
\section{Extraction}
\label{sec:examples}
We go in this section through a series of case studies of various recursive
algorithms.  We show how they can be encoded in our framework, how can we do
program calculation techniques for optimisation, and how can they be extracted
to idiomatic OCaml code. Our examples are the Quicksort and Mergesort
algorithms (Section~\ref{sec:sorting}), dynammic programming and Knapsack
(Section~\ref{sec:knap}), and examples of the shortcut deforestation
optimisation in our framework (Section~\ref{sec:shortcut}).
%The steps are the following:
%\begin{enumerate}
%  \item Write a specification as a morphism \coq{f : A ~> B}.
%  \item Use Coq's generalised rewriting and the laws of morphisms to 
%    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
%  \item Project the function \coq{h : A -> B} from \coq{g}
%    (i.e. \coq{h = app g}).
%  \item Extract \coq{h} to OCaml.
%\end{enumerate}
%Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
%\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
%proofs from the code, and inline our definitions. To ease the application of
%these steps, we provide the following helper type:
%\begin{coqcode}
%  Record Ext (f : A ~> B) := 
%    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
%\end{coqcode}
%Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
%since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
%will be the function \coq{target} that is extensionally equal to \coq{f}.  We
%use a shortcut tactic, \coq{calculate}, that is a synonym to 
%\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
%\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
%complete the definition with \coq{reflexivity}, and this will instantiate the
%metavariable \coq{?Goal0} to our desired function. We will use this extensively
%in our examples.
%\mvol{This is a very dry overview that doesn't need to be here. We can just jump
%  right into the examples.}
%
%\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
%  be used for a little overview of the rest of the section, like: ``We will go over a series
%  of case studies of various recursive algorithms, showing how they can be encoded
%  in our framework and extracted to idiomatic OCaml code.''}
%
%\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
%  Each of these is written like a case study, with a particular program worked out all the way
%  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
%  a sentence at the beginning of each section re-iterating this to make the whole thing
%  flow more smoothly.}
